An Accurate Model for SARS-CoV-2 pooled RT-PCR Test Errors

### 2021-03-31 comment.tex
commitz

\begin{abstract}
\textbf{Background}: Pooling is a popular strategy for increasing
SARS-CoV-2 testing throughput. One popular pooling scheme is Dorfman
pooling: test $N$ individuals simultaneously. If the test is positive
--- retest each individual separately. However, requiring more than
one positive test may lead to increased false-negative rates.

\textbf{Methods}: We analyze the false-negative rate (i.e., the
probability of a negative result for an infected individual) of
Dorfman pooling via a new probabilistic model. We demonstrate that
different, previously made probabilistic assumptions regarding pooling
are unlikely in light of empiric data. Our model is conservative in
that it ignores sample dilution effects, which can only worsen pooling
performance.

\textbf{Results}: We show that one can expect a 60-80\% increase in
false-negative rates under Dorfman pooling, for reasonable parameter
values. Moreover, we show that the false-negative rates under Dorfman
pooling increase when the prevalence of infection decreases.

\textbf{Discussion}: In most pooling schemes, identifying an infected
individual requires positive results in multiple tests and hence
substantially increases false-negative rates. Furthermore, this
phenomenon is more pronounced when infection prevalence is low ---
exactly when pooling is most efficient. Thus, pooling presents an
inherent trade-off: it is most efficient when it is least
accurate. The deterioration of false-negative rates and the
aforementioned trade-off are inherent problems of pooling schemes and
should be kept in mind by practitioners and policy makers.
\end{abstract}
\newpage

@@

We calculate the probability that Donald is mistakenly identified as
not infected, henceforth referred to as the scheme's false-negative
rate and denoted $\mi$. In order to correctly identify Donald
as infected, both pooled and separate tests have to yield a positive
result. Thus, the scheme's false-negative rate is:
\begin{align}\label{eq:sfn}
    \begin{split}
        \mi :&= 1 - \Se\Pr(\data=1|\theta=1)\\
        %
        &= 1 - \Se\left [1 - \Sp(1-\Se)(1-q\Se))^{N-1}\right].
    \end{split}
\end{align}

@@

Interestingly, an increase in infection prevalence monotonically
decreases the scheme false-negative rate, as can also be easily seen
from \eqref{eq:sfn}. For the chosen parameter ranges, the increase in
the single test false-negative rates increases the relative error
$\err{exact}$. These effects can be seen in Figure \ref{fig1} (left
panel), upon conditioning on pool size. Extending the range for $\Sp$
yields no qualitative differences. We further compare $\err{exact}$ to
$\err{common}$, showing the discrepancy changes as a function of both
prevalence and the single test sensitivity (Figure \ref{fig1}, right
panel).
\begin{figure}[H]
  \centering
  \includegraphics[width=\textwidth]{heatmap_sfn.jpg}
  \caption{Relative increase in Dorfman pooling false-negative rates
    $\err{exact}$. Left: Colors represent $\err{exact}$, the relative
    percentage increase in the scheme false-negative rates relative
    to the single test false-negative rates (eq
    \eqref{eq:erel}). Right: colors represent the difference between
    $\err{common}$ and $\err{exact}$. The disease prevalence $q$, is
    varied on the x-axis, while the test sensitivity is varied on
    the y-axis. Pool size $N$, was chosen according to $q$ as in
    \cite{DorfmanYuvalDor}.}\label{fig1}
\end{figure}

\section*{Discussion}
%\subsection{Short recap}
Here, we presented a novel probabilistic model for the outcomes of
pooled PCR tests, parametrized for SARS-CoV-2. Contrary to the common
assumption, we assume, based on data \cite{Salazar}, that multiple
infected individuals increase the likelihood of a positive pooled test
result. A direct consequence of our model is that false-negative rates
depend on infection prevalence. Specifically, low values of infection
prevalence increase $\err{exact}$. These results remain qualitatively
similar under varying parameter values, in the observed ranges
\cite{KitComparison,EstimatingRatesKucrika, EstimatingRatesLourenco,
  InterpretingCOVID19Test} (Figure \ref{fig1}).
Our results lead to a conflict: low infection prevalence leads to high
efficiency of Dorfman pooling \cite{DorfmanYuvalDor}, while also
increasing false-negative rates.

@@

Thus, finding improved pooling schemes is imperative. An especially
important consideration in designing such schemes is explicitly taking
the intrinsic PCR error rates into account. We \cite{DOPE}, as well as
others \cite{BayesianDorfman}, have developed such models, although their
implementation in real-world settings has not yet commenced.

>>>>

\begin{abstract}
pooling is a popular strategy for increasing
SARS-CoV-2 testing throughput. One popular pooling scheme is Dorfman
pooling: test $N$ individuals simultaneously. If the test is positive, each individual is  then tested separately. When error rates are taken into consideration,
most analyses assume that including extra infected samples in a pooled test has no effect 
on the pooled test outcome. We refute this assumption with experimental data
and suggest a novel probabilistic model for the outcomes of pooled tests. Our model is
parsimonous and assumes as little as possible. We analyze the false-negative
rate (i.e., the probability of a negative result for an infected individual) of
Dorfman pooling via a our model. We show that one can expect a 60-80\% increase in
false-negative rates under Dorfman pooling, for reasonable parameter
values. Moreover, we show that the false-negative rates under Dorfman
pooling increase when the prevalence of infection decreases.
In most pooling schemes, identifying an infected
individual requires positive results in multiple tests and hence
substantially increases false-negative rates. Furthermore, this
phenomenon is more pronounced when infection prevalence is low ---
exactly when pooling is most efficient. Thus, pooling presents an
inherent trade-off: it is most efficient when it is least
accurate. The deterioration of false-negative rates and the
aforementioned trade-off are inherent problems of pooling schemes and
should be kept in mind by practitioners and policy makers.
\end{abstract}
\newpage

@@

We calculate the probability that Donald is mistakenly identified as
not infected, henceforth referred to as the scheme's false-negative
rate and denoted $\mi$. In order to correctly identify an infected individual
as infected, both pooled and separate tests have to yield a positive
result. Thus, the scheme's false-negative rate is:
\begin{align}\label{eq:sfn}
    \begin{split}
        \mi :&= 1 - \Se\Pr(\data=1|\theta=1)\\
        %
        &= 1 - \Se\left [1 - \Sp(1-\Se)(1-q\Se))^{N-1}\right].
    \end{split}
\end{align}

@@

Interestingly, an increase in infection prevalence monotonically
decreases the scheme false-negative rate, as can also be easily seen
from \eqref{eq:sfn}. For the chosen parameter ranges, the increase in
the single test false-negative rates increases the relative error
$\err{exact}$. These effects can be seen in Figure~\ref{fig1} (left
panel), upon conditioning on pool size. Extending the range for $\Sp$
yields no qualitative differences. We further compare $\err{exact}$ to
$\err{common}$, showing the discrepancy changes as a function of both
prevalence and the single test sensitivity (Figure~\ref{fig1}, right
panel).
\begin{figure}[H]
  \centering
  \includegraphics[width=\textwidth]{heatmap_sfn.jpg}
  \caption{Relative increase in Dorfman pooling false-negative rates
    $\err{exact}$. Left: Colors represent $\err{exact}$, the relative
    percentage increase in the scheme false-negative rates relative
    to the single test false-negative rates (eq
    \eqref{eq:erel}). Right: colors represent the difference between
    $\err{common}$ and $\err{exact}$. The disease prevalence $q$, is
    varied on the x-axis, while the test sensitivity is varied on
    the y-axis. Pool size $N$, was chosen according to $q$ as in
    \cite{DorfmanYuvalDor}.}\label{fig1}
\end{figure}

\section*{Discussion}
%\subsection{Short recap}
Here, we presented a novel probabilistic model for the outcomes of
pooled PCR tests, parametrized for SARS-CoV-2. Contrary to the common
assumption, we assume, based on data (Table~\ref{table}), that multiple
infected individuals increase the likelihood of a positive pooled test
result. A direct consequence of our model is that false-negative rates
depend on infection prevalence. Specifically, low values of infection
prevalence increase $\err{exact}$. These results remain qualitatively
similar under varying parameter values, in the observed ranges
\cite{KitComparison,EstimatingRatesKucrika, EstimatingRatesLourenco,
  InterpretingCOVID19Test} (Figure \ref{fig1}).
Our results lead to a conflict: low infection prevalence leads to high
efficiency of Dorfman pooling \cite{DorfmanYuvalDor}, while also
increasing false-negative rates.

@@

Thus, improving pooling schemes is imperative. An especially
important consideration in designing such schemes is explicitly taking
the intrinsic PCR error rates into account. We \cite{DOPE}, as well as
others \cite{BayesianDorfman}, have developed such models, although they have not yet been implemented in real-world settings.
